---
title: "CLB Business Licensing Trends Analysis"
output:
  html_notebook: default
  pdf_document: default
---

## Prerequisites 

Make sure you have [tidyverse](http://tidyverse.org/), [lubridate](https://github.com/hadley/lubridate), and [prophet](https://facebookincubator.github.io/prophet/) libraries installed in your R enviornment before we begin.

```{r}
library(tidyverse)
library(lubridate)
library(prophet)
```

I also import some commonly used functions and code for analyses.

```{r}
source("../R/utils.R")
```

# Analysis request from management

> "Anything about seasonal [business licensing] fluctuation? For instance, the licenses in the latter half of the year generally take longer because there is a lot of construction going on."

We will be importing the latest Business Licensing dataset available from City of Long Beach / Financial management and first analyze trends for Active licenses. This analysis will include all licensing activity (new, issued, renewed).

Afterward, we will see if there is any time series trends for how many days it takes to issue a license to a new applicant. 

## Import the business licensing dataset

```{r}
business_licenses <- read_csv("../data/raw/business_license.csv",
                              col_types = cols(
                                servicedate = col_date("%Y-%m-%d %H:%M:%S"),
                                startdate = col_date("%Y-%m-%d %H:%M:%S"),
                                milestonedate = col_date("%Y-%m-%d %H:%M:%S"),
                                load_date = col_datetime(),
                                bid_code = col_character(),
                                licensecat = col_character(),
                                zip = col_character()
                              ))
```

## Data cleanup

### Filtering out Inactive licenses

As per Bretty Yakus, an **Active** business license is defined by Financial Management (FM) as business licenses that have a status of:

  * Active
  * CollcInBus
  * Expired (a licensee might become reactivated or go into collections)
  * Pending
  * Conditionl

Thereby, **Inactive** licenses are those that are not of the above statuses.

Create a new dataframe for "Active" licenses.

```{r}

active_licstatus = c('Active', 'CollcInBus', 'Expired', 'Pending', 'Reactivate', 'Conditionl')
active_milestones = c('Issued', 'Renewed', 'Collections', 'Pre-Collections', 'Pre-Renew', 'Renewed')

active_licenses <- business_licenses %>%
  filter(licstatus %in% active_licstatus & milestone %in% active_milestones)
```

### Get a daily count of Active/Issued licenses

Clean up steps required to count our data: 

1. In order to get a set of reliable data, we need to get a daily summary of Active/Issued business licenses. Note that we are filtering out all licenses before **August 2015**. This is due to the new version of the LMR/Hansen module foe business licenses going online August 2015. Prior to this system upgrade milestones weren't track electronically for applications - thus not having an electonic log to check when a license was first issued.
2. We need to log-transforming our y variable in order to more clearly see increases in variation over time. This is especially important for exponentiall growing trends.
3. Finally, we need rename our date/milestone date and count variable to **ds** and **y** respectively. The `prophet` library expects these named parameters to create our forecasting model.

```{r}

analysis_start_date <- as_date("2015-08-01")

active_licenses_daily_count <- active_licenses %>%
  mutate(ds = milestonedate) %>%
  filter(ds >= analysis_start_date) %>%
  group_by(ds) %>% 
  summarise(
    y = log(n())
  ) %>% 
  select(ds, y)
```

We are using the **milestonedate** in our analysis. The milestone date is used by Hansen as the log date/time stamp to capture capture events for a business licensee.

## Forecasting

### Create the model

We fit the model by passing in the historical dataframe and instantiating a new `Prophet` object.

```{r}
m <- prophet(active_licenses_daily_count)
```

### Predicting growth

Predictions are then made on a dataframe with a column named `ds` containing the dates for which predictions are to be made. The `make_future_dataframe` function takes the model object and a number of periods to forecast and produces a suitable dataframe. By default it will also include the historical dates so we can evaluate in-sample fit.

We will now make predictions with the model for the next 12 months. We are passing in the montly freuency paramater to account for non-daily data as business licenses aren't processed daily. By sticking to monthly forecasts, we can reduce some of the noise in our data. 

```{r}
future <- make_future_dataframe(m, periods = 12, freq = 'm')
forecast <- predict(m, future)
```

The forecast can be visualized using the `plot` function.

```{r}
plot_forecast(m,
              forecast,
              xlabel = 'Year',
              ylabel = "Log Count")
ggsave("../figs/business_licensing_forecast.png")
```


We can use the `prophet_plot_components` function to see the forecast components. By default you'll see the trend, yearly, seasonality, and weekly seasonality of the time series.

```{r}
prophet_plot_components(m, forecast)
```

#### Trend forecast component

```{r}
prophet_plot_trend_component(m, forecast, xlabel = 'Year', ylabel = 'Trend')
ggsave("../figs/business_licensing_forecast_trend_component.png")
```

#### Yearly forecast component

```{r}
prophet_plot_yearly_component(m, forecast, xlabel = 'Day of year', ylabel = 'Log Count')
ggsave("../figs/business_licensing_forecast_year_component.png")
```

## Export data

Export our processed dataframes for archival and sharing purposes.

```{r}
write_csv(active_licenses, "../data/interim/active_business_licenses.csv")
write_csv(active_licenses_daily_count, "../data/processed/active_business_licenses_count.csv")
write_csv(forecast, "../data/processed/business_licensing_forecast.csv")
```

# How about Days to Issue?

## Data import

```{r}

```


## Data cleanup


